Interactive web-based flood modeling at country wide scale and planter size resolution.

### 2014-05-16 article.tex
fix reviewer 1 and fix header layout to line up with word based pdf

% extra packages not in the template
% Don't color hyperlinks

\usepackage{wrapfig}
\usepackage{paralist}
\usepackage[draft]{hyperref}
\usepackage{amsmath}
\usepackage[nolist]{acronym}
\usepackage{sansmath}
\usepackage{listings}
\usepackage[list=true]{subcaption}

@@

\pdfpagewidth=210 true mm
\pdfpageheight=297 true mm

@@

\section{Introduction}
The flooding of rural and urban areas is an increasing hazard to society. Accurate and timely predictions are essential for the water manager to be able to carefully plan, design and control the effects of flooding hazard.
For the prediction of these flooding hazards requires a numerical model that represents the physical processes (rain, evaporation, flow, infiltration). This model is fed with measurements from current or historic events. The results of these calculations, the expected flooding, is used for damage estimates, dike design, spatial and emergency planning and for coordinating response efforts. Several challenges required an innovative redesign of both the numerical model and the information infrastructure where it is nested within.

@@

The amount of data involved in preparing the model is rapidly increasing. There is an increase in supply of data caused by the ever growing measurement resolution and spatial coverage. For example the topology of waterbodies at a spatial coverage of a small country are used to determine if the water will find its way into ones neighborhood. The topography and infiltration rate at a spatial resolution of a planter determine if the water will flow to ones backyard or the neighbors. How can we design a platform that is suitable for this new modeling approach and meets the corresponding challenges?

The \threedi  platform, in which these approaches and several new ones are applied, consists of a new overland flooding model (1D/2D) and a corresponding infrastructure that has been applied for interactive web-based modeling of rural and urban areas at coverages of \orderof(1000km2) at a resolution of \orderof(0.1m). Here we describe the advances in the numerical methods, architecture and visualization techniques.

@@

By introspection, the outside world (e.g., a Python-based website, a C\#-\ac{GUI}, a MATLAB-script) can examine model state at run time. More specifically: get current water levels and flow fields, current running state of water pumps and other structures, active meteorological forcings. Each of these state variables can just as easily be \emph{set} as they are \emph{get}. These functions are implemented in correspondence with the \ac{BMI} standard \citep{Peckham2013}. The \ac{API} exposes memory pointers to relevant model state variables. This is how the model directly `feels' and uses its new state.

@@

The heterogeneous landscape and land use are specified in the model as gigapixel bitmaps for soil types, crop types, drainage resistance and maximum infiltration.  Infiltration, evaporation and drainage are evaluated at pixel resolution, whereas the computational grid cells are larger and ensure high performance.

@@

To enable such an environment the most logical choice is to use a web-based environment. Over the last years this choice has become more apparent. The large list of browser enhancements (for example inline \ac{SVG}, canvas, plugin free video \citep{Berjon2014} and \ac{CSS} animations \citep{Jackson2013}) allow for rich visualizations independent of the device type. An overview of the system architecture is given in \autoref{fig:arch}. The architecture follows the general ``Model View Controller'' pattern \citep{Gamma1994}, with caching of events and data to accommodate for the model continuously running in the background. The implementation of this system required several technical innovations and social considerations.

@@

One of the differences when comparing a desktop user to a web-based user is the short attention span. If no indication of progress is given for longer than two second users will start reloading, clicking erratically or will just close the website, searching for a new ``frisson'' \citep{Carr2011}. To make sure that the model is ready to run when a user arrives a lot of attention is given to the provisioning of the model. Provisioning, in this context, is preparing a model controller so that it starts its computation on request. Machines are prepared and put on hold at one of the \ac{IaaS} providers (such as Amazon). These machines are prepared with a container that contains the model schematization, tables and grids and hot restart files. These files are read into memory so that the model is ready to start its computation when needed, resulting in a time between a user logging in and a model start in under 5 seconds.

@@

Another performance technique used to achieve the high framerate is to use the \emph{Mahlen nach Zahlen} method, a reference to a product from the German company Ravensburger~\textregistered. A map with cell numbers is prepared in advance. Once a quantity for a new timestep becomes available the map is filled in with the colors corresponding to the cell numbers. This achieves much faster rendering times than drawing the cells using geometry drawing function.

@@

Another planned step is to incorporate or preferably to couple the system with other physical models. The monolithic approach of putting all processes into one model has proven cumbersome in previous implementations. The current model controller can easily incorporate other models because the communication and control are already out of the model engine. Water quality, coastal and socio-economic models are possible candidates for inclusion in the platform.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{iemss}
\bibliography{./bibliography}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{acronym}[AAAAA]
\acro{CSS}{Cascading Style Sheets}
\acro{MMI}{Model Message Interface}
\acro{JSON}{JavaScript Object Notation}
\acro{0MQ}[\O{}MQ]{Zero Message Queue}
\acro{SVG}{Scalable Vector Graphics}
\acro{H2O}[$\mathrm{H_2O}$]{water}
\acro{BMI}{Basic Model Interface}
\acro{API}{Application Programming Interface}
\acro{GUI}{Graphical User Interface}
\acro{DEM}{Digital Elevation Model}
\acro{HTTP}{Hyper Text Transfer Protocol}
\acro{IaaS}{Infrastructure as a Service}
\end{acronym}

>>>>

% extra packages not in the template
% Don't color hyperlinks
\pdfpagewidth=210 true mm
\pdfpageheight=297 true mm

\usepackage{wrapfig}
\usepackage{paralist}
\usepackage[draft]{hyperref}
\usepackage{xcolor}
% \hypersetup{
%   colorlinks=true,
%   citecolor=black,
%   filecolor=black,
%   linkcolor=black,
%   urlcolor=blue
% }
\urlstyle{same}
\usepackage{amsmath}
\usepackage[nolist]{acronym}
\usepackage{sansmath}
\usepackage{listings}
\usepackage[list=true]{subcaption}

@@

@@

\section{Introduction}
The flooding of rural and urban areas is an increasing hazard to society. Accurate and timely predictions are essential for the water manager to be able to carefully plan, design, and control the effects of flooding hazard.
For the prediction of these flooding hazards requires a numerical model that represents the physical processes (rain, evaporation, flow, infiltration). This model is fed with measurements from current or historic events. The results of these calculations, the expected flooding, is used for damage estimates, dike design, spatial and emergency planning, and for coordinating response efforts. Several challenges required an innovative redesign of both the numerical model and the information infrastructure where it is nested within.

@@

The amount of data involved in preparing the model is rapidly increasing. There is an increase in supply of data caused by the ever growing measurement resolution and spatial coverage. For example, the topology of waterbodies at a spatial coverage of a small country is used to determine if the water will find its way into one's neighborhood. The topography and infiltration rate at a spatial resolution of a planter determine if the water will flow to one's backyard or the neighbors. How can we design a platform that is suitable for this new modeling approach and meets the corresponding challenges?

The \threedi  platform, in which these approaches and several new ones are applied, consists of a new overland flooding model (1D/2D) and a corresponding infrastructure that has been applied for interactive web-based modeling of rural and urban areas at coverages of \orderof(1000km2) at a resolution of \orderof(0.1m). Here we describe the advances in the numerical methods, architecture, and visualization techniques.

@@

By introspection, the outside world (e.g., a Python-based website, a C\#-\ac{GUI}, a MATLAB-script) can examine model state at run time. More specifically: get current water levels and flow fields, current running state of water pumps, and other structures, active meteorological forcings.
% Fix sentence
 Each of these state variables can just as easily be \emph{set} as they are \emph{get}. These functions are implemented in correspondence with the \ac{BMI} standard \citep{Peckham2013}. The \ac{API} exposes memory pointers to relevant model state variables. This is how the model directly `feels' and uses its new state.

@@

The heterogeneous landscape and land use are specified in the model as gigapixel bitmaps for soil types, crop types, drainage resistance, and maximum infiltration.  Infiltration, evaporation, and drainage are evaluated at pixel resolution, whereas the computational grid cells are larger and ensure high performance.

@@

To enable such an environment the most logical choice is to use a web-based environment. Over the last years this choice has become more apparent. The large list of browser enhancements (for example inline \ac{SVG}, canvas, plugin free video \citep{Berjon2014}, and \ac{CSS} animations \citep{Jackson2013}) allow for rich visualizations independent of the device type. An overview of the system architecture is given in \autoref{fig:arch}. The architecture follows the general ``Model View Controller'' pattern \citep{Gamma1994}, with caching of events and data to accommodate for the model continuously running in the background. The implementation of this system required several technical innovations and social considerations.

@@

One of the differences when comparing a desktop user to a web-based user is the short attention span. If no indication of progress is given for longer than two second users will start reloading, clicking erratically or will just close the website, searching for a new ``frisson'' \citep{Carr2011}. To make sure that the model is ready to run when a user arrives a lot of attention is given to the provisioning of the model. Provisioning, in this context, is preparing a model controller so that it starts its computation on request. Machines are prepared and put on hold at one of the \ac{IaaS} providers (such as Amazon). These machines are prepared with a container that contains the model schematization, tables and grids, and hot restart files. These files are read into memory so that the model is ready to start its computation when needed, resulting in a time between a user logging in and a model start in under 5 seconds.

@@

Another performance technique used to achieve the high framerate is to use the ``Mahlen nach Zahlen'' method, a reference to a product from the German company Ravensburger~\textregistered. A map with cell numbers is prepared in advance. Once a quantity for a new timestep becomes available the map is filled in with the colors corresponding to the cell numbers. This achieves much faster rendering times than drawing the cells using geometry drawing function.

@@

Another planned step is to incorporate or preferably to couple the system with other physical models. The monolithic approach of putting all processes into one model has proven cumbersome in previous implementations. The current model controller can easily incorporate other models because the communication and control are already out of the model engine. Water quality, coastal, and socio-economic models are possible candidates for inclusion in the platform.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{apalike}
\bibliography{./bibliography}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{acronym}[AAAAA]
\acro{CSS}{Cascading Style Sheets}
\acro{MMI}{Model Message Interface}
\acro{JSON}{JavaScript Object Notation}
\acro{0MQ}[\O{}MQ]{Zero Message Queue}
\acro{SVG}{Scalable Vector Graphics}
\acro{H2O}[$\mathrm{H_2O}$]{water}
\acro{BMI}{Basic Model Interface}
\acro{API}{Application Programming Interface}
\acro{GUI}{Graphical User Interface}
\acro{DEM}{Digital Elevation Model}
\acro{HTTP}{Hyper Text Transfer Protocol}
\acro{IaaS}{Infrastructure as a Service}
\end{acronym}
